---
title: "Data Exploration for WASH Status at Refugee Camps"
author:
  name: "Shannon Teri"
  orcid: "0009-0000-9889-1427"
  email: "shannon.teri@colorado.edu"
  affiliation:
    name: "University of Colorado Boulder"
    department: "Department of Civil, Environmental and Architectural Engineering"
    city: "Boulder"
    state: "CO"
    country: "USA"
date: 2026-07-01
format: html
editor: visual
execute:
  echo: false
  output: false
  warning: false
  message: false
---

# Introduction

Looking at the conditions of refugee camps across 36 countries and 227 locations. The UNHCR Information Management System (IRIS) collected the data as an aggregate of conditions across specfic camps. I want to use this data to visualize these conditions and make cross-comparisons of similarities and differences based on location, time, and reason.

# Methods

## Reading the Data

```{r}
library(tidyverse)
library(here)
library(gt)
library(countrycode)

unhcr_wash_data <- read_csv(here::here("data/raw/refugee_camp_water_data.csv"))
```

## Initial Data Tidying

Initial Looks

```{r}
# view the first few rows
head(unhcr_wash_data)

# check the number of rows/columns
dim(unhcr_wash_data)

# get the header names
colnames(unhcr_wash_data)

# number of inputs per country, checking that the values equal number of inputs in data
country_check <- unhcr_wash_data %>% 
  count(country)
sum(country_check$n)
# checks! = 6425
```

Clean the Data

```{r}
# drop rows that don't have substantial data
wash_data_lim <- unhcr_wash_data %>% 
  mutate(toilets_per_HH_percentage = coalesce(households_with_toilet, household_toilet)) %>% 
  mutate(region = countrycode(country,
         origin = "country.name",
         destination = "un.regionsub.name")) |> 
  mutate(region = factor(region, levels = c("Latin America and the Caribbean",
                                            "Northern Africa",
                                            "Sub-Saharan Africa",
                                            "Western Asia",
                                            "Southern Asia"))) |> 
  rename(liters_person_day = liters_per_person_per_day) |> 
  filter(!if_all(persons_per_handpump:reporting_annual, is.na)) %>% 
  select(-c(form_id, 
            location_id, 
            reporting_monthly, 
            reporting_annual, 
            household_toilet, 
            households_with_toilet)) %>% 
  relocate(refugee_pop, .before = persons_per_handpump) %>% 
  relocate(c(persons_per_shower, 
             persons_per_hygiene_promoter, 
             persons_per_toilet), .before = liters_person_day) %>% 
  relocate(liters_per_person_household, .after = liters_person_day) %>% 
  relocate(c(defecate_in_toilet, 
             toilets_per_HH_percentage, 
             solid_waste_disposal_access, 
             menstrual_hygiene_satisfaction, 
             access_to_soap), .after = persons_per_toilet) %>% 
  rename(safe_water_perc = chlorinated_safe_water_quality,
         liters_per_HH_per_day = liters_per_person_household,
         storage10L_HH_perc = potable_water_storage_10l,
         safe_water_collect_HH_perc = protected_water_sources,
         defecate_in_toilet_HH_perc = defecate_in_toilet,
         access_to_soap_HH_perc = access_to_soap,
         solid_waste_disposal_HH_perc = solid_waste_disposal_access) %>% 
  relocate(safe_water_perc, .before = safe_water_collect_HH_perc) |> 
  relocate(c(region, country), .before = location_name)

# get a new year column so for better summary
wash_data_lim <- wash_data_lim |> 
  mutate(year = year(start_date),
         month = month(start_date + (end_date - start_date) / 2))

wash_data_lim <- wash_data_lim |> 
  mutate(report_date = sprintf("%d-%02d", year, month)) |> 
  relocate(report_date, .before = start_date)
# some are a bit weird, but when checking against later dataset, they are excluded

wash_data_lim <- wash_data_lim |> 
  select(-c(year, month))
```

Cleaned Data Summaries

```{r}
# summarize by year, but using wash_data_lim to get all countries/locations, yyyy-mm
wash_data_timeline_lim_full <- wash_data_lim |> 
  group_by(report_date) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) |> 
  pivot_longer(
    cols = c(country_count, location_count),
    names_to = "Measure", 
    values_to = "Count"
  ) |> 
  pivot_wider(
    names_from = report_date,
    values_from = Count
  )

# summarize by year, but using wash_data_lim to get all countries/locations, yyyy
wash_data_timeline_lim <- wash_data_lim |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(year) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) |> 
  pivot_longer(
    cols = c(country_count, location_count),
    names_to = "Measure", 
    values_to = "Count"
  ) |> 
  pivot_wider(
    names_from = year,
    values_from = Count
  )
```

Core Data Tidy

```{r}
# get rid of any that are over 3215 cells (half)
# get rid of columns that only have 6 cells worth of data
wash_data_core <- wash_data_lim |> 
  select(-c(persons_per_handpump, 
            defecate_in_toilet_HH_perc, 
            solid_waste_disposal_HH_perc,
            menstrual_hygiene_satisfaction, 
            access_to_soap_HH_perc, 
            liters_per_HH_per_day, 
            non_chlorinated_0_cfu, 
            storage10L_HH_perc, 
            safe_water_collect_HH_perc))

# figure out what data sets aren't viable (not enough information)
wash_data_core <- wash_data_core |> 
  mutate(not_NA = rowSums(!is.na(wash_data_core))) |> 
  filter(not_NA > 6) |> 
  relocate(c(liters_person_day, safe_water_perc), .before = persons_per_shower) |> 
  relocate(persons_per_hygiene_promoter, .after = toilets_per_HH_percentage)

# get rid of the not_NA column (not needed)
wash_data_core <- wash_data_core |> 
  select(-c(not_NA, start_date, end_date))


```

Core Summaries

```{r}
# summarize by county
wash_data_core_summary <- wash_data_core %>% 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(country) %>% 
  summarise(mean_pop = mean(refugee_pop, na.rm = TRUE),
            sum_cities = n_distinct(location_name),
            sum_reports = n(),
            sum_report_years = n_distinct(year),
            post_emergency = mean((post_emergency == "Post-emergency"), na.rm = TRUE) * 100)
# anything not "post-emergency" means that it was active during an emergency

```

```{r}
#| output: TRUE
#| label: data_reported
#| fig-cap: "Table 1: Summary that Averages the WASH Data Collected by Country"
#| fig-alt: "A table that lists the average statistics on the population, number, reporting period, and percentage of post-emergency of refugee camps by country."
#| fig-width: 12

wash_data_core_summary |> 
  gt() |> 
  tab_header(
    title = "Refugee Camps by Country",
    subtitle = "Data Reported bewteen 2013 and 2024"
  ) |> 
  fmt_number(
    columns = mean_pop,
    decimals = 2
  ) |> 
  fmt_number(
    columns = post_emergency,
    decimals = 2,
    suffix = "%"
  ) |> 
  cols_label(
    country = "Country",
    mean_pop = "Avg Refugee Camp Population",
    sum_cities = "Number of Refugee Camps",
    sum_reports = "Number of Reports",
    sum_report_years = "Number of Years with Reports",
    post_emergency = "% Refugee Camps Post-Emergency"
  )
```

```{r}
# SUMMARIES OF REPORTING PERIODS BY COUNTRY AND LOCATION (basically showing activity)
# summarize by year, but only using core for countries/locations being used, yyyy
wash_data_timeline <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(year, region) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) 

wash_data_timeline_yr <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(year) |> 
  summarise(country_count = n_distinct(country))
```

Initial Visualization

```{r}
#| output: TRUE
#| label: reports_per_year
#| fig-cap: "Figure 1: Number of Refugee Camps Reported per Year"
#| fig-alt: "Bar chart of the number of refugee camps reported on each year. This is cross-referenced with a line graph that shows the change in the number of countries reporting on refugee camps each year. The bar chart is further identified by color to indicate how many refugee camps are present per region."
#| fig-width: 9

ggplot() +
  geom_col(data = wash_data_timeline,
           mapping = aes(x = year,
                         y = location_count,
                         fill = region)) +
    scale_fill_brewer(palette = "Set1") +
  geom_point(data = wash_data_timeline_yr,
             mapping = aes(x = year, 
                           y= country_count,
                           group = 1)) +  
  geom_line(data = wash_data_timeline_yr,
            mapping = aes(x = year,
                          y = country_count,
                          linetype = "Countries Reporting",
                          group = 1)) +
  scale_y_continuous(
    name = "Number of Refugee Camps", 
    sec.axis = sec_axis(
      ~ .,
        name = "Number of Countries Reporting"
      )
    ) +
  labs(title = "Number of Refugee Camps Reported per Year",
       subtitle = "Compared to Number of Countries Reporting per Year",
       x = "Year", 
       y = "Number of Refugee Camps",
       fill = "Regions",
       linetype = "Countries")
```

Write CSV

```{r}
write_csv(wash_data_lim,
          here::here("data/processed/refugee_camp_water_data_tidied.csv"))
```

# Results

According the UNHCR:

```{r}
# FOCUS SPECIFICALLY ON SUB-SAHARHAN AFRICA SINCE IT HAS THE MOST DATA
# Starting with laying out the standards and actuals per year

core_SSA <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  filter(region == "Sub-Saharan Africa", 
         year >= 2018, 
         liters_person_day < 75,
         persons_per_shower < 300, 
         persons_per_tap < 750,
         persons_per_toilet < 100) |> 
  group_by(region, country, year) |> 
  summarise(liters_person_day = mean(liters_person_day, na.rm = TRUE),
            persons_per_tap = mean(persons_per_tap, na.rm = TRUE),
            persons_per_toilet = mean(persons_per_toilet, na.rm = TRUE),
            persons_per_shower = mean(persons_per_shower, na.rm = TRUE)) |> 
  pivot_longer(
    cols = c(liters_person_day,
             persons_per_tap,
             persons_per_toilet,
             persons_per_shower),
    names_to = "Indicators", 
    values_to = "Values"
  )

reference <- tibble(
  Indicators = c("liters_person_day",
                 "persons_per_tap",
                 "persons_per_toilet",
                 "persons_per_shower"),
  Baseline = c(20,
             100,
             20,
             50)
)

core_SSA <- left_join(core_SSA, reference, by = join_by("Indicators"))

```

```{r}
#| output: TRUE
#| label: subset_data_SSA
#| fig-cap: "Figure 2: Averages of Indicators by Country per Year"
#| fig-alt: "Four line graphs that outline the average values of each indicator by country per year. This is referenced against the international standard that should be met at refugee sites to maintain proper WASH infrastructure."
#| fig-width: 9
#| fig-height: 9

ggplot(data = core_SSA,
       mapping = aes(x = year, 
                     y = Values, 
                     group = country, 
                     color = country)) +
  geom_point() +
  geom_line() +
  geom_hline(
    aes(yintercept = Baseline),
    linetype = "dashed",
    linewidth = 1.5,
    color = "red"
  ) +
  facet_wrap(~Indicators,
             scales = "free_y",
             labeller = labeller(
               Indicators = c(
                 "liters_person_day" = "Liters per Person per Day",
                 "persons_per_shower" = "Persons per Shower",
                 "persons_per_tap" = "Persons per Tap",
                 "persons_per_toilet" = "Persons per Toilet"
               )
             )) +
  scale_color_viridis_d() +
  labs(title = "Average of Indicators by Country",
       subtitle = "Compared against International Standard",
       x = "Year", 
       y = "Average per Indicator", 
       color = "Country") +
  theme_minimal()
```

```{r}
# preparing a summary table of how often refugee camps don't meet international standard
avg_summary <- core_SSA |> 
  pivot_wider(
    names_from = "Indicators", 
    values_from = "Values"
  ) |> 
  group_by(country) |> 
  summarise(avg_liters = mean(liters_person_day, na.rm = TRUE),
            avg_taps = mean(persons_per_tap, na.rm = TRUE),
            avg_toilets = mean(persons_per_toilet, na.rm = TRUE),
            avg_shower = mean(persons_per_shower, na.rm = TRUE)) 

```

```{r}
results_fin <- tibble(
  Indicators = c("Liters per Person per Day",
                 "Persons per Tap", 
                 "Persons per Toilet",
                 "Persons per Shower"),
  Results = c(mean((avg_summary$avg_liters) < 20)*100,
              mean((avg_summary$avg_taps) > 100) *100, 
              mean((avg_summary$avg_toilets) > 20) *100, 
              mean((avg_summary$avg_shower) > 50) * 100)
)
```

```{r}
core_SSA_location <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  filter(region == "Sub-Saharan Africa", 
         year >= 2018, 
         liters_person_day < 75,
         persons_per_shower < 200, 
         persons_per_tap < 750,
         persons_per_toilet < 100, 
         refugee_pop < 150000) |> 
  group_by(country, location_name) |> 
  summarise(liters_person_day = mean(liters_person_day, na.rm = TRUE),
            persons_per_tap = mean(persons_per_tap, na.rm = TRUE),
            persons_per_toilet = mean(persons_per_toilet, na.rm = TRUE),
            persons_per_shower = mean(persons_per_shower, na.rm = TRUE),
            refugee_pop = mean(refugee_pop, na.rm = TRUE)) |>   
  pivot_longer(
    cols = c(liters_person_day,
             persons_per_tap,
             persons_per_toilet,
             persons_per_shower),
    names_to = "Indicators", 
    values_to = "Values"
  )

core_SSA_location <- left_join(core_SSA_location, reference, by = join_by("Indicators"))
```

```{r}
#| output: TRUE
#| label: subset_data_SSA_location
#| fig-cap: "Figure 3: Refugee Camp Population and Relation to WASH Standards"
#| fig-alt: "Four line graphs that outline the average values of each indicator by location, further categorized by country by color. This is referenced against the international standard that should be met at refugee sites to maintain proper WASH infrastructure as well as an established trendline comparing population to resource management."
#| fig-width: 9
#| fig-height: 9

ggplot(data = core_SSA_location,
       mapping = aes(x = refugee_pop, 
                     y = Values, 
                     color = country)) +
  geom_point(size=2) +
  geom_smooth(
    aes(group = 1, 
        linetype = "Trend"),
    method = "lm", 
    se = FALSE,
    linewidth = 0.5) +
  geom_hline(
    aes(yintercept = Baseline, 
        linetype = "Baseline"),
    linetype = "dashed",
    linewidth = 1,
    color = "red"    
  ) +
  facet_wrap(~Indicators,
             scales = "free_y",
             labeller = labeller(
               Indicators = c(
                 "liters_person_day" = "Liters per Person per Day",
                 "persons_per_shower" = "Persons per Shower",
                 "persons_per_tap" = "Persons per Tap",
                 "persons_per_toilet" = "Persons per Toilet"
               )
             )) +
  scale_color_viridis_d() +
  labs(title = "Comparison of Population to WASH Resource Management",
       subtitle = "Averages of Each Refugee Camp, further Identified by Country",
       x = "Refugee Population", 
       y = "Average Value per Indicator", 
       color = "Country", 
       linetype = "Data Trend"
  ) +
  theme_minimal()
```

```{r}
#| output: TRUE
#| label: indicator_summary_table
#| fig-cap: "Table 2: Average Percentage of Refugee Camps that do not meet WASH International Standards"
#| fig-alt: "A table that lists the WASH indicators and a percentage that represents how many do not meet the international standards."
#| fig-width: 9

results_fin |> 
  gt() |> 
  tab_header(
    title = "Percentage of Refugee Camps above International Standard",
    subtitle = "Determined by Indicator"
  ) |> 
  fmt_number(
    columns = Results,
    decimals = 2) |> 
  cols_label(
    Indicators = "Indicator",
    Results = "Results",
  )
```

# Conclusions

## Summary of Findings

## Questions and Next Steps

Right now I'm working in generalizations by region, I would want to do more of a deep dive into the specifics per Refugee Camp. It would also be interesting to see the their locations, how they coorelate, and get more information about why they were formed.
