---
title: "Data Exploration for WASH Status at Refugee Camps"
author:
  name: "Shannon Teri"
  orcid: "0009-0000-9889-1427"
  email: "shannon.teri@colorado.edu"
  affiliation:
    name: "University of Colorado Boulder"
    department: "Department of Civil, Environmental and Architectural Engineering"
    city: "Boulder"
    state: "CO"
    country: "USA"
date: 2026-07-01
format: html
editor: visual
---

# Introduction

Looking at the conditions of refugee camps across 36 countries and 227 locations. The UNHCR Information Management System (IRIS) collected the data as an aggregate of conditions across specfic camps. I want to use this data to visualize these conditions and make cross-comparisons of similarities and differences based on location, time, and reason.

# Methods

```{r}
library(dplyr)
library(tidyverse)
library(here)

unhcr_wash_data <- read_csv(here::here("data/raw/refugee_camp_water_data.csv"))
```

## Reading the Data

## Initial Data Tidying

```{r}
# view the first few rows
head(unhcr_wash_data)

# check the number of rows/columns
dim(unhcr_wash_data)

# get the header names
colnames(unhcr_wash_data)

# number of inputs per country, checking that the values equal number of inputs in data
country_check <- unhcr_wash_data %>% 
  count(country)
sum(country_check$n)
# checks! = 6425

wash_data_combined <- unhcr_wash_data %>% 
  mutate(toilets_per_household_per = coalesce(households_with_toilet, household_toilet)) 

# drop rows that don't have substantial data
wash_data_lim <- unhcr_wash_data %>% 
  mutate(toilets_per_HH_percentage = coalesce(households_with_toilet, household_toilet)) %>% 
  filter(!if_all(persons_per_handpump:reporting_annual, is.na)) %>% 
  select(-c(form_id, location_id, reporting_monthly, reporting_annual, household_toilet, households_with_toilet)) %>% 
  relocate(refugee_pop, .before = persons_per_handpump) %>% 
  relocate(c(persons_per_shower, persons_per_hygiene_promoter, persons_per_toilet), .before = liters_per_person_per_day) %>% 
  relocate(liters_per_person_household, .after = liters_per_person_per_day) %>% 
  relocate(c(defecate_in_toilet, toilets_per_HH_percentage, solid_waste_disposal_access, menstrual_hygiene_satisfaction, access_to_soap), .after = persons_per_toilet) %>% 
  rename(safe_water_perc = chlorinated_safe_water_quality,
         liters_per_HH_per_day = liters_per_person_household,
         storage10L_HH_perc = potable_water_storage_10l,
         safe_water_collect_HH_perc = protected_water_sources,
         defectae_in_toilet_HH_perc = defecate_in_toilet,
         access_to_soap_HH_perc = access_to_soap,
         solid_waste_disposal_HH_perc = solid_waste_disposal_access) %>% 
  relocate(safe_water_perc, .before = safe_water_collect_HH_perc)


#check <- wash_data_lim %>% 
#  mutate(
#    status = case_when(
#      !is.na(household_toilet) & !is.na(households_with_toilet) ~ "Both",
#      !is.na(household_toilet) | !is.na(households_with_toilet) ~ "One",
#      TRUE ~ "Neither"
#    )
#  ) 
  
#check %>% 
#  count(status)

# summarize
wash_data_lim %>% 
  group_by(country) %>% 
  summarise(mean_pop = mean(refugee_pop, na.rm = TRUE))

#I'm having issues with summarise

write_csv(wash_data_lim,
          here::here("data/processed/refugee_camp_water_data_tidied.csv"))

            
```

# Results

# Conclusions

## Summary of Findings

## Questions and Next Steps
