---
title: "Data Exploration for WASH Status at Refugee Camps"
author:
  name: "Shannon Teri"
  orcid: "0009-0000-9889-1427"
  email: "shannon.teri@colorado.edu"
  affiliation:
    name: "University of Colorado Boulder"
    department: "Department of Civil, Environmental and Architectural Engineering"
    city: "Boulder"
    state: "CO"
    country: "USA"
date: 2026-07-01
format: html
editor: visual
---

# Introduction

Looking at the conditions of refugee camps across 36 countries and 227 locations. The UNHCR Information Management System (IRIS) collected the data as an aggregate of conditions across specfic camps. I want to use this data to visualize these conditions and make cross-comparisons of similarities and differences based on location, time, and reason.

# Methods

## Reading the Data

```{r}
library(tidyverse)
library(here)
library(gt)
library(countrycode)

unhcr_wash_data <- read_csv(here::here("data/raw/refugee_camp_water_data.csv"))
```

## Initial Data Tidying

Initial Looks

```{r}
# view the first few rows
head(unhcr_wash_data)

# check the number of rows/columns
dim(unhcr_wash_data)

# get the header names
colnames(unhcr_wash_data)

# number of inputs per country, checking that the values equal number of inputs in data
country_check <- unhcr_wash_data %>% 
  count(country)
sum(country_check$n)
# checks! = 6425
```

Clean the Data

```{r}
# drop rows that don't have substantial data
wash_data_lim <- unhcr_wash_data %>% 
  mutate(toilets_per_HH_percentage = coalesce(households_with_toilet, household_toilet)) %>% 
  mutate(region = countrycode(country,
         origin = "country.name",
         destination = "un.regionsub.name")) |> 
  mutate(region = factor(region, levels = c("Latin America and the Caribbean",
                                            "Northern Africa",
                                            "Sub-Saharan Africa",
                                            "Western Asia",
                                            "Southern Asia"))) |> 
  filter(!if_all(persons_per_handpump:reporting_annual, is.na)) %>% 
  select(-c(form_id, 
            location_id, 
            reporting_monthly, 
            reporting_annual, 
            household_toilet, 
            households_with_toilet)) %>% 
  relocate(refugee_pop, .before = persons_per_handpump) %>% 
  relocate(c(persons_per_shower, 
             persons_per_hygiene_promoter, 
             persons_per_toilet), .before = liters_per_person_per_day) %>% 
  relocate(liters_per_person_household, .after = liters_per_person_per_day) %>% 
  relocate(c(defecate_in_toilet, 
             toilets_per_HH_percentage, 
             solid_waste_disposal_access, 
             menstrual_hygiene_satisfaction, 
             access_to_soap), .after = persons_per_toilet) %>% 
  rename(safe_water_perc = chlorinated_safe_water_quality,
         liters_per_HH_per_day = liters_per_person_household,
         storage10L_HH_perc = potable_water_storage_10l,
         safe_water_collect_HH_perc = protected_water_sources,
         defecate_in_toilet_HH_perc = defecate_in_toilet,
         access_to_soap_HH_perc = access_to_soap,
         solid_waste_disposal_HH_perc = solid_waste_disposal_access) %>% 
  relocate(safe_water_perc, .before = safe_water_collect_HH_perc) |> 
  relocate(c(region, country), .before = location_name)

# get a new year column so for better summary
wash_data_lim <- wash_data_lim |> 
  mutate(year = year(start_date),
         month = month(start_date + (end_date - start_date) / 2))

wash_data_lim <- wash_data_lim |> 
  mutate(report_date = sprintf("%d-%02d", year, month)) |> 
  relocate(report_date, .before = start_date)
# some are a bit weird, but when checking against later dataset, they are excluded

wash_data_lim <- wash_data_lim |> 
  select(-c(year, month))
```

Cleaned Data Summaries

```{r}
# summarize by year, but using wash_data_lim to get all countries/locations, yyyy-mm
wash_data_timeline_lim_full <- wash_data_lim |> 
  group_by(report_date) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) |> 
  pivot_longer(
    cols = c(country_count, location_count),
    names_to = "Measure", 
    values_to = "Count"
  ) |> 
  pivot_wider(
    names_from = report_date,
    values_from = Count
  )

# summarize by year, but using wash_data_lim to get all countries/locations, yyyy
wash_data_timeline_lim <- wash_data_lim |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(year) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) |> 
  pivot_longer(
    cols = c(country_count, location_count),
    names_to = "Measure", 
    values_to = "Count"
  ) |> 
  pivot_wider(
    names_from = year,
    values_from = Count
  )
```

Core Data Tidy

```{r}
# get rid of any that are over 3215 cells (half)
# get rid of columns that only have 6 cells worth of data
wash_data_core <- wash_data_lim |> 
  select(-c(persons_per_handpump, 
            defecate_in_toilet_HH_perc, 
            solid_waste_disposal_HH_perc,
            menstrual_hygiene_satisfaction, 
            access_to_soap_HH_perc, 
            liters_per_HH_per_day, 
            non_chlorinated_0_cfu, 
            storage10L_HH_perc, 
            safe_water_collect_HH_perc))

# figure out what data sets aren't viable (not enough information)
wash_data_core <- wash_data_core |> 
  mutate(not_NA = rowSums(!is.na(wash_data_core))) |> 
  filter(not_NA > 6) |> 
  relocate(c(liters_per_person_per_day, safe_water_perc), .before = persons_per_shower) |> 
  relocate(persons_per_hygiene_promoter, .after = toilets_per_HH_percentage)

# get rid of the not_NA column (not needed)
wash_data_core <- wash_data_core |> 
  select(-c(not_NA, start_date, end_date))


```

Core Summaries

```{r}
# summarize by county
wash_data_core_summary <- wash_data_core %>% 
  group_by(country) %>% 
  summarise(mean_pop = mean(refugee_pop, na.rm = TRUE),
            sum_cities = n_distinct(location_name),
            sum_reports = n(),
            sum_report_years = n_distinct(year(start_date)),
            post_emergency = mean((post_emergency == "Post-emergency") * 100))
# anything not "post-emergency" means that it was active during an emergency

wash_data_core_summary |> 
  gt() |> 
  tab_header(
    title = "Refugee Camps by Country",
    subtitle = "Data Reported bewteen 2013 and 2024"
  ) |> 
  fmt_number(
    columns = mean_pop,
    decimals = 2
  ) |> 
  fmt_number(
    columns = post_emergency,
    decimals = 2,
    suffix = "%"
  ) |> 
  cols_label(
    country = "Country",
    mean_pop = "Avg Refugee Camp Population",
    sum_cities = "Number of Refugee Camps",
    sum_reports = "Number of Reports",
    sum_report_years = "Number of Years with Reports",
    post_emergency = "% Refugee Camps Post-Emergency"
  )

# SUMMARIES OF REPORTING PERIODS BY COUNTRY AND LOCATION (basically showing activity)
# summarize by year, but only using core for countries/locations being used, yyyy
wash_data_timeline <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(year, region) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) # |> 
  #pivot_longer(
   # cols = c(country_count, location_count),
   # names_to = "Measure",
  #  values_to = "Count"
  #) |> 
  #pivot_wider(
   # names_from = year,
    #values_from = Count
  #)

# summarize by year, but only using core for countries/locations being used, yyyy-mm
wash_data_timeline_full <- wash_data_core |> 
  group_by(report_date, region) |> 
  summarise(country_count = n_distinct(country),
            location_count = n_distinct(location_name)) #|> 
#  pivot_longer(
 #   cols = c(country_count, location_count),
  #  names_to = "Measure",
   # values_to = "Count"
#  ) |> 
 # pivot_wider(
  #  names_from = report_date,
   # values_from = Count
#  )
```

Initial Visualization

```{r}
#| label: country_by_year
#| fig-cap: "Figure 1: Number of Countries with Reported Refugee Camps per Year"
#| fig-alt: "Bar chart that of the number of countries with reportings per year. Further identified by color for distribution of countries by global region."

ggplot(data = wash_data_timeline,
       mapping = aes(x = year,
                     y = country_count,
                     fill = region)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Countries with Refugee Camp Reports per Year", 
       subtitle = "Identified by Region",
       x = "Year",
       y = "Number of Countries",
       fill = "Region")

#| label: location_by_year
#| fig-cap: "Figure 2: Number of Refugee Camps Reported per Year"
#| fig-alt: "Bar chart that of the number of refugee camp reportings per year. Further identified by color for distribution of countries by global region."

ggplot(data = wash_data_timeline,
       mapping = aes(x = year,
                     y = location_count,
                     fill = region)) +
  geom_col() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Number of Refugee Camps Reported per Year", 
       subtitle = "Identified by Region",
       x = "Year",
       y = "Number of Refugee Camps",
       fill = "Region")

```

Write CSV

```{r}
write_csv(wash_data_lim,
          here::here("data/processed/refugee_camp_water_data_tidied.csv"))
```

# Results

According the UNHCR:

Water Supply

- 20 liters / person / day (by country)

```{r}
liters20_region <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(region, year) |> 
  summarise(liters_per_person_per_day = mean(liters_per_person_per_day, na.rm = TRUE)) |> 
  filter(liters_per_person_per_day != 0)

#| label: country_by_year
#| fig-cap: "Figure 1: Number of Countries with Reported Refugee Camps per Year"
#| fig-alt: "Bar chart that of the number of countries with reportings per year. Further identified by color for distribution of countries by global region."

ggplot(data = liters20_region,
       mapping = aes(x = year,
                     y = liters_per_person_per_day, 
                     color = region,
                     group = region)) +
  geom_line() + 
  geom_point() +
  scale_color_brewer(palette = "Set2") + 
  labs(
    title = "Liters per Person per Day",
    subtitle = "Identified by Region", 
    x = "Year", 
    y = "Liters per Person per Day", 
    color = "Region"
  )

```

- by camp

```{r}
liters20_country <- wash_data_core |> 
  mutate(year = substr(report_date, "1", "4")) |> 
  group_by(country, year) |> 
  summarise(liters_per_person_per_day = mean(liters_per_person_per_day, na.rm = TRUE)) |>
  filter(liters_per_person_per_day != 0)
  # takes out the ones that don't have data (listed at 0)

#| label: country_by_year
#| fig-cap: "Figure 1: Number of Countries with Reported Refugee Camps per Year"
#| fig-alt: "Bar chart that of the number of countries with reportings per year. Further identified by color for distribution of countries by global region."

ggplot(data = liters20_country,
       mapping = aes(x = year,
                     y = liters_per_person_per_day, 
                     color = country,
                     group = country)) +
  geom_line() + 
  geom_point() +
  scale_color_brewer(palette = "Set2") + 
  labs(
    title = "Liters per Person per Day",
    subtitle = "Identified by Region", 
    x = "Year", 
    y = "Liters per Person per Day", 
    color = "Country"
  )
```

- serve max 100 people (by county)

```{r}
wash_data_core |> 
  group_by(country) |> 
  summarise(persons_per_tap = mean(persons_per_tap, na.rm = TRUE)) |> 
  filter(!is.na(persons_per_tap))
# takes away Brazil that didn't have info on this
```

- serve max 100 people (by camp)

```{r}
wash_data_core |> 
  group_by(country, location_name) |> 
  summarise(persons_per_tap = mean(persons_per_tap, na.rm = TRUE)) |> 
  filter(persons_per_tap != 0)
```

Bathrooms:

- 1 bathroom / 20 people

```{r}
wash_data_core |> 
  group_by(country) |> 
  summarise(persons_per_toilet = mean(persons_per_toilet, na.rm = TRUE)) |> 
  filter(persons_per_toilet != 0)
```

- by camp

```{r}
wash_data_core |> 
  group_by(country, location_name) |> 
  summarise(persons_per_toilet = mean(persons_per_toilet, na.rm = TRUE)) |> 
  filter(persons_per_toilet != 0)
```

- 1 shower / 50 people (country)

```{r}
wash_data_core |> 
  group_by(country) |> 
  summarise(persons_per_shower = mean(persons_per_shower, na.rm = TRUE)) |> 
  filter(persons_per_shower != 0)
```

- by camp

```{r}
wash_data_core |> 
  group_by(country, location_name) |> 
  summarise(persons_per_shower = mean(persons_per_shower, na.rm = TRUE)) |> 
  filter(persons_per_shower != 0)
```

Population:

```{r}
# chart with the mean refugee pop per country
wash_data_core |> 
  group_by(country) |> 
  summarise(refugee_pop = mean(refugee_pop, na.rm = TRUE)) |> 
  filter(refugee_pop != 0)
```

- 20,000 people per camp

```{r}
wash_data_core |> 
  group_by(country, location_name) |> 
  summarise(refugee_pop = mean(refugee_pop, na.rm = TRUE)) |> 
  filter(refugee_pop != 0)
```

# Conclusions

## Summary of Findings

## Questions and Next Steps
