---
title: "Data Exploration for WASH Status at Refugee Camps"
author:
  name: "Shannon Teri"
  orcid: "0009-0000-9889-1427"
  email: "shannon.teri@colorado.edu"
  affiliation:
    name: "University of Colorado Boulder"
    department: "Department of Civil, Environmental and Architectural Engineering"
    city: "Boulder"
    state: "CO"
    country: "USA"
date: 2026-07-01
format: html
editor: visual
---

# Introduction

Looking at the conditions of refugee camps across 36 countries and 227 locations. The UNHCR Information Management System (IRIS) collected the data as an aggregate of conditions across specfic camps. I want to use this data to visualize these conditions and make cross-comparisons of similarities and differences based on location, time, and reason.

# Methods

```{r}
library(tidyverse)
library(here)

unhcr_wash_data <- read_csv(here::here("data/raw/refugee_camp_water_data.csv"))
```

## Reading the Data

## Initial Data Tidying

```{r}
# view the first few rows
head(unhcr_wash_data)

# check the number of rows/columns
dim(unhcr_wash_data)

# get the header names
colnames(unhcr_wash_data)

# number of inputs per country, checking that the values equal number of inputs in data
country_check <- unhcr_wash_data %>% 
  count(country)
sum(country_check$n)
# checks! = 6425

# drop rows that don't have substantial data
wash_data_lim <- unhcr_wash_data %>% 
  mutate(toilets_per_HH_percentage = coalesce(households_with_toilet, household_toilet)) %>% 
  filter(!if_all(persons_per_handpump:reporting_annual, is.na)) %>% 
  select(-c(form_id, 
            location_id, 
            reporting_monthly, 
            reporting_annual, 
            household_toilet, 
            households_with_toilet)) %>% 
  relocate(refugee_pop, .before = persons_per_handpump) %>% 
  relocate(c(persons_per_shower, 
             persons_per_hygiene_promoter, 
             persons_per_toilet), .before = liters_per_person_per_day) %>% 
  relocate(liters_per_person_household, .after = liters_per_person_per_day) %>% 
  relocate(c(defecate_in_toilet, 
             toilets_per_HH_percentage, 
             solid_waste_disposal_access, 
             menstrual_hygiene_satisfaction, 
             access_to_soap), .after = persons_per_toilet) %>% 
  rename(safe_water_perc = chlorinated_safe_water_quality,
         liters_per_HH_per_day = liters_per_person_household,
         storage10L_HH_perc = potable_water_storage_10l,
         safe_water_collect_HH_perc = protected_water_sources,
         defecate_in_toilet_HH_perc = defecate_in_toilet,
         access_to_soap_HH_perc = access_to_soap,
         solid_waste_disposal_HH_perc = solid_waste_disposal_access) %>% 
  relocate(safe_water_perc, .before = safe_water_collect_HH_perc)

# get a new year column so for better summary
wash_data_lim <- wash_data_lim |> 
  mutate(year = year(start_date),
         month = month(start_date + (end_date - start_date) / 2))

wash_data_lim <- wash_data_lim |> 
  mutate(report_date = sprintf("%d-%02d", year, month)) |> 
  relocate(report_date, .before = start_date)
# some are a bit weird, but when checking against later dataset, they are excluded

wash_data_lim <- wash_data_lim |> 
  select(-c(year, month))

# check size of dataset
dim(wash_data_lim)

# check how many empty cells per column
colSums(is.na(wash_data_lim))

# get rid of any that are over 3215 cells (half)
# get rid of columns that only have 6 cells worth of data
wash_data_core <- wash_data_lim |> 
  select(-c(persons_per_handpump, 
            defecate_in_toilet_HH_perc, 
            solid_waste_disposal_HH_perc,
            menstrual_hygiene_satisfaction, 
            access_to_soap_HH_perc, 
            liters_per_HH_per_day, 
            non_chlorinated_0_cfu, 
            storage10L_HH_perc, 
            safe_water_collect_HH_perc))

# figure out what data sets aren't viable (not enough information)
wash_data_core <- wash_data_core |> 
  mutate(not_NA = rowSums(!is.na(wash_data_core))) |> 
  filter(not_NA > 6)

# get rid of the not_NA column (not needed)
wash_data_core <- wash_data_core |> 
  select(-not_NA)

# summarize by county
wash_data_core %>% 
  group_by(country) %>% 
  summarise(mean_pop = mean(refugee_pop, na.rm = TRUE),
            sum_instances = n(),
            sum_cities = n_distinct(location_name),
            sum_report_years = n_distinct(year(start_date)))

write_csv(wash_data_lim,
          here::here("data/processed/refugee_camp_water_data_tidied.csv"))

            
```

# Results

# Conclusions

## Summary of Findings

## Questions and Next Steps
